---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Data Visualization Project: U.S Domestic Airline Carriers 2003 - 2016
#### Summary
To investigate and produce a data visualization of the flight performance of
US airline carriers over the years 2003 to 2016.  The data documented the
carriers performance in terms of metrics such as number of flights, arrivals per
airport, delays and flights cancelled.  This information was downloaded from 
RITA at: http://www.transtats.bts.gov/OT_Delay/OT_DelayCause1.asp

### Design

**Exploratory Data Analysis**

I downloaded the dataset as a csv file for exploratory data analysis (EDA) in R 
studio.  I thought originally it would be more interesting to focus on airports
such as asking the question "which airport has the most flight arrivals?" but
after looking at the data the information was not so interesting as the busiest
airports were at obvious geographical hubs such as Hartsfield-Jackson Atlanta 
and Chicago O'Hara.

I therefore started to look at the actual carriers and ask question such as 
"Which carrier has the most flights?" and "Which carrier has the least delays?".
After looking at the dataset there were many carriers in operation (27 in total).
I realized that to display this information graphically would be difficult so I
decided to cut down the number of carriers to five.  The five biggest carriers 
were defined by the summing the number of flights per carrier and taking the 
five carriers with greatest totals:   

```{r code1,  include=FALSE}
require(knitr)
require(markdown)
library(dplyr)
library(ggplot2)
setwd('/home/riched/udacity/nanodegree/P6')
airlines <- read.csv('airline_delay_causes_best.csv')
airlines = airlines[is.na(airlines$arr_flights) == 0,]
str(airlines)

airlines <-rename(airlines, weather_ct = X.weather_ct,arr_delay = X.arr_delay,
                  carrier_delay=X.carrier_delay, month=X.month)
str(airlines)
summary(airlines)

best5 = tail(sort(tapply(airlines$arr_flights, airlines$carrier_name, sum)),n=5)
best5.carrier_name = labels(best5)
best5.carrier_name = best5.carrier_name[[1]]
best5_airlines = airlines[airlines$carrier_name %in% best5.carrier_name  ,]
best5_airlines$X = NULL
best5_airlines$carrier_name = factor(best5_airlines$carrier_name)
levels(best5_airlines$carrier_name)

summary_data = aggregate(best5_airlines[,7:ncol(best5_airlines)],
                            by = list(best5_airlines$year,best5_airlines$carrier_name),
                            FUN=sum, na.rm=TRUE)
summary_data <-rename(summary_data, year = Group.1,carrier_name = Group.2)

summary_data$avg_arr_del15 = 100 * summary_data$arr_del15/summary_data$arr_flights
summary_data$avg_carrier_ct = 100 * summary_data$carrier_ct/summary_data$arr_flights
summary_data$avg_ontime = 100 - summary_data$avg_arr_del15
summary_data$num_flights = summary_data$arr_flights
```

```{r code2,echo=FALSE}
levels(best5_airlines$carrier_name)
```

The dataset was processed to produce another csv file **summary_airline.csv**
.  After EDA I decided to use the data visualization to
answer the question "Amongst the bigger carriers is there a relationship between
carrier size and flight delay times?".  Because of the flexibility of R compared
to processing the data using dimple, I decided to generate plots first in R to
get a feel of how they would look using **dimple.js**.  The R code is included
in the file **airlines.r**.

The first plot shows the actual number of flights per year per 
carrier over the time period.  Clearly Southwest Airlines has the most flights
over this period:  

```{r code3, echo=FALSE}
ggplot(data = summary_data,
       aes(x = year, y = num_flights)) +
  geom_line(aes(color = carrier_name))
```

The second plot shows the percentage of on-time flights per carrier, we can see
that Delta airlines significantly improved its number of on-time flights.  This
number is defined as arriving within 15 mins of due arrival time:

```{r code4, echo=FALSE}

ggplot(data = summary_data,
       aes(x = year, y = avg_ontime)) +
  geom_line(aes(color = carrier_name))
```

The final plot shows the actual percentage of late flights caused by the actual 
carrier as opposed to other delay causes such as weather and security issues.
We can see that at the beginning of the period Skywest Airlines has much poorer
performance than the other airlines but improved greatly, to the the extent of
being the best in this category:

```{r code5, echo=FALSE}
ggplot(data = summary_data,
       aes(x = year, y = avg_carrier_ct)) +
  geom_line(aes(color = carrier_name))

```

**Data Visualisation**

I implemented the visualization using **dimple.js**.  Time was of the essence to
me,
so I only used a small amount of **d3.js** code.  I decided to employ three 
charts similar to the ones I completed in R studio. After the initial 
visualization I made some changes to the visualization:

* I overlayed a scatter plot to the series lines
* I added a title for each of the plots

After making these changes to the visualization, I asked three people for 
feedback:

* 'the figures should be placed with the number of flights per carrier at the
top on its own line'
* 'the colors for the plots are too bright make them more subtle'
* 'the placement of the figures dos not look right'

After reviewing the feedback, I placed the Number of flights per carrier plot
at the top of the page in the center.  I then placed the remaining two plots in
the center directly under the first plot.  I researched colors and found some
colors that i felt to be less bright and more easy on the eye.

The final visualization is available as **airlines_3.html**. From the
visualization
it is clear that Southwest Airlines has the most flights but this does not
translate into delay times.  In the percentage of on time flights Delta Airlines 
performs best and in the percentage of late flights caused by the actual carrier
performance amongst the carriers is very similar.  Southwest Airlines does not
provide better time performance for its flights, there must be other reasons
it has many more flights than the other carriers.

#### References
* http://www.color-hex.com/color-names.html
* http://dimplejs.org/
* http://stackoverflow.com/
* https://github.com/d3/d3-3.x-api-reference/blob/master/API-Reference.md




