---
output: word_document
---

### Red Wine Exploratory Data Analysis


#### 1 Introduction

This data set contains 1,599 red wines with 11 variables on the chemical 
properties of the wine. At least 3 wine experts rated the quality of each wine,
providing a rating between 0 (very bad) and 10 (very excellent).

Guiding Question: Which chemical properties influence the quality of red wines?

```{r message=FALSE, warning=FALSE,echo = FALSE}

library(scales)
library(memisc)
library(reshape)
library(gridExtra)
library(ggplot2)
library(dplyr)
library(GGally)
library(grid)

#Load in the data
setwd('/home/riched/udacity/nanodegree/P4/project')
wines <- read.csv('wineQualityReds.csv')
```

#### 2 Summary of Data
Obtain a summary of the data set:
```{r Summary, message=FALSE, warning=FALSE}
#str(wines)
summary(wines)
```

* There are 1599 observations of 13 variables
* Quality ranges from a value of 3 to 8 with a mean value of 5.6.  From research
  these values came from median ratings of three wine experts which would tend 
  to smooth the data values leading to less outliers.
* The variables are continuous apart from X (observation identifier) and those 
  that have a .sulfur.dioxide suffix
  
```{r message=FALSE, warning=FALSE}
#Clean up and format
wines <- subset(wines, select = - X)

```

#### 3 Univarate Summary Plot
General overview of the variables in figure form.
```{r message=FALSE, warning=FALSE,fig.width=10, fig.height=6,echo=FALSE}
#Summary of variables, adjust binwidths to get clearer picture

p1<- qplot(data = wines,x=wines$fixed.acidity,binwidth = 0.2)
p2<- qplot(data = wines,x=wines$volatile.acidity,binwidth = 0.02)
p3<- qplot(data = wines,x=wines$citric.acid,binwidth = 0.05)
p4<- qplot(data = wines,x=wines$residual.sugar,binwidth = 0.5)
p5<- qplot(data = wines,x=wines$chlorides,binwidth = 0.05)
p6<- qplot(data = wines,x=wines$free.sulfur.dioxide,binwidth = 1)
p7<- qplot(data = wines,x=wines$total.sulfur.dioxide,binwidth = 1)
p8<- qplot(data = wines,x=wines$density,binwidth = 0.0005)
p9<- qplot(data = wines,x=wines$pH,binwidth = 0.05)
p10<-qplot(data = wines,x=wines$sulphates,binwidth = 0.1)
p11<-qplot(data = wines,x=wines$alcohol,binwidth = 0.5)
p12<-ggplot(wines, aes(x=as.factor(quality))) + geom_bar()

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12,ncol=4)

```

Residual.sugar, chlorides, free.sulfur.dioxide, total.sulfur.dioxide
and sulphates appear to have exterme outliers.  The Figure can be replicated 
with these outliers hidden for clearer plotting purposes.

```{r message=FALSE, warning=FALSE,fig.width=10, fig.height=6,echo=FALSE}
#Summary of variables, remove outliers to get clearer picture
p1<- qplot(data = wines,x=wines$fixed.acidity,binwidth = 0.2)
p2<- qplot(data = wines,x=wines$volatile.acidity,binwidth = 0.02)
p3<- qplot(data = wines,x=wines$citric.acid,binwidth = 0.05)
p4<- qplot(data = wines,x=wines$residual.sugar,binwidth = 0.5) + xlim(0,4)
p5<- qplot(data = wines,x=wines$chlorides,binwidth = 0.025)+xlim(0,0.15)
p6<- qplot(data = wines,x=wines$free.sulfur.dioxide,binwidth = 1)+xlim(0,60)
p7<- qplot(data = wines,x=wines$total.sulfur.dioxide,binwidth = 1)+xlim(0,200)
p8<- qplot(data = wines,x=wines$density,binwidth = 0.0005)
p9<- qplot(data = wines,x=wines$pH,binwidth = 0.05)
p10<-qplot(data = wines,x=wines$sulphates,binwidth = 0.1)+xlim(0,1.5)
p11<-qplot(data = wines,x=wines$alcohol,binwidth = 0.5)
p12<-ggplot(wines, aes(x=as.factor(quality))) + geom_bar()

grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12,ncol=4)

```

#### 4 Univarate Analysis

Density and pH have little variance and appear to have normal distribution 
curves.  The dioxide based values, fixed and volatile acidity, sulphates and
residual sugar are positive skewed.  Sugar and chlorides have are long tailed 
and thus have extreme outliers.  The distribution of the quality ranking are
narrow, there are few extreme rankings, most of the rankings are in the middle
of the distribution (ratings of 5 and 6).    

Create another categorical  variable classifying wines as 'Bad', 'Average' or
'Good':
```{r}
wines$rating<-cut(as.numeric(wines$quality), c(2.5,4.5,6.5,8.5),
                   labels=c('bad','average','good'))
```

Fixed acidity is slightly skewed to higher acidity but with a tail extending to 
a value of 15.9 $\frac{g}{dm^{3}}$. The mean 7.90$\frac{g}{dm^{3}}$ and median 
8.32 $\frac{g}{dm^{3}}$ values were quite similar demonstrating the lack of 
outliers in the distribution.  Volatile acidity had a similar distribution to 
fixed, plotting on a log scale made both appear more normal distributed.  

```{r analysis, message=FALSE, warning=FALSE,echo=FALSE}
#Summary of variables, log scale shown for comparison
p1<-ggplot(data = wines,
       aes(x = fixed.acidity)) +
  geom_histogram() + scale_x_log10()+xlab('fixed.acidity log scale')
p2<-ggplot(data = wines,
       aes(x = volatile.acidity)) +
  geom_histogram() + scale_x_log10()+xlab('volatile.acidity log scale')
grid.arrange(p1,p2,ncol=2)

```

Superimposing the boxplot onto the scatter plots of these acidity values is 
a good way to summarise the data.  From the scatterplot/boxplot
volatile.acidity clearly has a more normal distribution as the boxplot
is approximately at the centre of the points on the scatterplot.

```{r message=FALSE, warning=FALSE,echo=FALSE}

grid.arrange(ggplot(wines, aes( x = 1, y = fixed.acidity ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = fixed.acidity  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

grid.arrange(ggplot(wines, aes( x = 1, y = volatile.acidity ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = volatile.acidity  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)


```

```{r message=FALSE, warning=FALSE}
#summary of fixed acidity
summary(wines$fixed.acidity)

```

```{r message=FALSE, warning=FALSE}
#summary of fixed volatile acidity
summary(wines$volatile.acidity)

```

Citric acid had a median value of 0.260.  Citric acid did not show a normal 
distribution when plotted on log scale.The most common citric acid value for the
wines appears around zero, to be more exact:

```{r, message=FALSE, warning=FALSE,echo=FALSE}

grid.arrange(ggplot(wines, aes( x = 1, y = citric.acid ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = citric.acid  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

#percentage of zero values 
number_of_zero_values <-132/length(wines$citric.acid)
```

For citric acid, 132 wines had zero values,  this equates to 8% of the wines, 
considering the bin values surrounding the zero value are much smaller than 132,
this could be indicate a problem with the data collection.  

Residual sugar shows a concentration around the median of 2.2$\frac{g}{dm^{3}}$ 
with ouliers up to a max of 15.5$\frac{g}{dm^{3}}$. From research we would 
expect the great majority of wines between 1 and 4 $\frac{g}{dm^{3}}$ as red 
wines are not sweet.  The Boxplot is superimposed onto scatter plot to 
demonstrate the distribution of residal sugar values, the great majority of the
values in are in the 1 to 4 range.

```{r sugar, message=FALSE, warning=FALSE,echo=FALSE}

grid.arrange(ggplot(wines, aes( x = 1, y = residual.sugar ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = residual.sugar  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)
#table(wines$residual.sugar)
```

```{r, message=FALSE, warning=FALSE}
#summary of residual sugar
summary(wines$residual.sugar)
```

Chlorides had a normal distribution with a peak at 0.07 $\frac{g}{dm^{3}}$ and a 
long tail to 0.61 $\frac{g}{dm^{3}}$.  
```{r chlorides,message=FALSE, warning=FALSE,echo=FALSE}

grid.arrange(ggplot(wines, aes( x = 1, y = chlorides ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = chlorides  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

#table(wines$chlorides)
```

```{r, message=FALSE, warning=FALSE}
#summary of chlorides
summary(wines$chlorides)
```

Free and total sulfur dioxide are bunched at lower values.  Total sulfur dioxide
has a small number of extreme outliers. 

```{r diox,message=FALSE, warning=FALSE,echo=FALSE}

grid.arrange(ggplot(wines, aes( x = 1, y = free.sulfur.dioxide ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = free.sulfur.dioxide  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

grid.arrange(ggplot(wines, aes( x = 1, y = total.sulfur.dioxide ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = total.sulfur.dioxide  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)
```


```{r, message=FALSE, warning=FALSE}
#summary of free.sulfur.dioxide
summary(wines$free.sulfur.dioxide)
```

```{r, message=FALSE, warning=FALSE}
#summary of total.sulfur.dioxide
summary(wines$total.sulfur.dioxide)
```


From research free sulfur dioxide varies from about 40% to 75% of the total 
dioxide - before sulphates are added.  Lets add this ratio to data set:

```{r, message=FALSE, warning=FALSE}
 wines$free.sulfur.dioxide.ratio <- with( wines, free.sulfur.dioxide / total.sulfur.dioxide)
```

```{r diox2,message=FALSE, warning=FALSE,echo=FALSE}
ggplot(data = wines,
       aes(x = free.sulfur.dioxide.ratio)) +
  geom_histogram(binwidth = 0.0025) 
```

Reveals a high concentration at a ratio 0.5, maybe winemakers aim to achieve 
this ratio? 

Density and pH has a normal distribution with little variance.  

```{r density,message=FALSE, warning=FALSE,echo=FALSE}

grid.arrange(ggplot(wines, aes( x = 1, y = density ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = density  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)

grid.arrange(ggplot(wines, aes( x = 1, y = pH ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = pH  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)



```

```{r, message=FALSE, warning=FALSE}
#summary of density
summary(wines$density)
```

```{r, message=FALSE, warning=FALSE}
#summary of pH
summary(wines$pH)
```

Sulphates has a similar distribution as residual.sugar and chlorides
```{r sulphates,message=FALSE, warning=FALSE,echo=FALSE}

grid.arrange(ggplot(wines, aes( x = 1, y = sulphates ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = sulphates  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)
```

```{r, message=FALSE, warning=FALSE}
#summary of sulphates
summary(wines$sulphates)
```

Alcohol has a peak around 9.5, seems to be correlated with sulfur.dioxide. 
```{r alcohol, message=FALSE, warning=FALSE,echo=FALSE}

grid.arrange(ggplot(wines, aes( x = 1, y = alcohol ) ) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) ,
             ggplot(wines, aes( x   = alcohol  ) ) + 
                   geom_histogram(bins=30 ),ncol=2)
```

```{r, message=FALSE, warning=FALSE}
#summary of alcohol
summary(wines$alcohol)
```


#### 5 Univarate analysis

##### What is the structure of your dataset?
There are 1599 different observations of wines with 11 different wine variables.
All the variables were measured as floating numerical values except for the 
unique identifier X and the categorical variable quality.

##### What is/are the main feature(s) of interest in your dataset?
The main feature of interest is quality, how is the value of quality determined?

##### What other features in the dataset do you think will help support your 
#####  investigation into your feature(s) of interest?
All wines contain sulfur dioxide in various forms, sulphates are added by the 
winemaker as an additive.  For experienced tasters high concentrations of added
sulphates can be unpleasant.  Residual sugar will be lower in red wine but 
higher outliers may cause the wine to be too sweet resulting in lower quality
ratings. Acidity (fixed, volatile and citric) also plays a big part in wine 
quality, different combinations of these amounts will lead to a variation in the 
perceived quality.

#### Did you create any new variables from existing variables in the dataset?
I made new variables the ratio of free to total sulfur dioxide and a categorical
variable rating of quality.

#### Of the features you investigated, were there any unusual distributions? 
Did you perform any operations on the data to tidy, adjust, or change the form
of the data? If so, why did you do this?

The data was already in a tidy format, i did remove the variable X from the data
set as this variable simply represented the row index.  Citric acid stood out, 
8% of the wines had no citric acid.   Citric acid most commonly used in wine is
produced by the addition of acid supplements, to boost the wines total acidity, 
this suggests a significant proportion of the winemakers added no citric acid. 

#### 6 Bivarate Analysis
We can use a function to determine the correlation between the variables.
The pass each of the variables to the funcion in turn.

``` {r message=FALSE, warning=FALSE}
cor_test <- function(x, y) {
  return(cor.test(x, as.numeric(y))$estimate)
}
```


```{r corr, message=FALSE, warning=FALSE,fig='hide',echo=FALSE}

cor_test <- function(x, y) {
  return(cor.test(x, as.numeric(y))$estimate)
}

correlations <- c(
  cor_test(wines$fixed.acidity, wines$quality),
  cor_test(wines$volatile.acidity, wines$quality),
  cor_test(wines$citric.acid, wines$quality),
  cor_test(wines$residual.sugar, wines$quality),
  cor_test(wines$chlorides, wines$quality),
  cor_test(wines$free.sulfur.dioxide, wines$quality),
  cor_test(wines$total.sulfur.dioxide, wines$quality),
  cor_test(wines$density, wines$quality),
  cor_test(wines$pH, wines$quality),
  cor_test(wines$sulphates, wines$quality),
  cor_test(wines$alcohol, wines$quality))
names(correlations) <- c('fixed.acidity', 'volatile.acidity', 'citric.acid',
                         'residual.sugar',
                         'chlorides', 'free.sulfur.dioxide',
                         'total.sulfur.dioxide', 'density', 'pH',
                         'sulphates', 'alcohol')
correlations
```

There does not appear to be a high correlation value between variables in the 
data set and quality.  The strongest correlation with quality were alcohol 
(0.476), volatile.acidity (-0.391) sulphates (0.251) and citric acid (0.226).
The weakest correlations were residual sugar(0.014), free.sulfur.dioxide(-0.051)
and pf(-0.058).  

##### Alcohol v Quality
The median alcohol value vs quality shows an increase after a quality value of 5.
I suspect this may be due to the 'fuller' flavour as the alcohol content 
increases perceived as higher quality.  The trend line shows the realtionship
between quality and alcohol content.  

```{r corrplots1, message=FALSE, warning=FALSE,echo=FALSE}
with(wines, by(alcohol, factor(quality), summary))

ggplot(aes(factor(quality), 
            alcohol), 
        data = wines) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  geom_smooth(aes(quality-2, 
                    alcohol),
                method = "lm", 
                se = FALSE,size=2)

```

##### Volatile Acidity v Quality
Quality and volatile acidity displays a negative correlation, almost as strong a
correlation as alcohol vs quality but in the other direction.  Volatile.acidity 
appears to be an unwanted feature in wine, quality increases as it goes down.  
The lowest quality wines have a high median value of volatile acidity.  From 
research wine spoilage is legally defined by volatile acidity, largely composed 
of acetic acid.  Higher proportions of acetic acid also lead to unpleasant 
aromas in wine.  Many winemakers seek a low value of acetic acid as this adds 
to the perceived complexity of a wine.  These higher levels of volatile acidity 
(acetic acid) may therefore negatively impact on the taste and aroma of a wine.  

```{r message=FALSE, warning=FALSE,echo=FALSE}
with(wines, by(volatile.acidity, factor(quality), summary))

ggplot(aes(factor(quality), 
            volatile.acidity), 
        data = wines) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  geom_smooth(aes(quality-2, 
                    volatile.acidity),
                method = "lm", 
                se = FALSE,size=2)


```

##### Sulphates v Quality
The correlation between sulphates and quality was 0.25.  Not as high as alcohol 
and volatile acidity but there is a general trend for higher quality wines 
having higher sulphate values.  Sulphates are added by winemakers to prevent 
oxidation and keep unwanted bacteria at bay.  Therefore this is done not to 
improve the taste but to ensure the wine taste does not degrade, maybe the lower
quality wines have started to degrade affecting the perceived quality of the 
wine.  Wines witha quality of 5 or 6 had a large number of outliers which would
drive down the correlation value.  From the lowest quality factor the median
sulphate value increases until a quality factor of 7 and 8 where it stays the
same. 
```{r message=FALSE, warning=FALSE,echo=FALSE}
with(wines, by(sulphates, factor(quality), summary))

ggplot(aes(factor(quality), 
            sulphates), 
        data = wines) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  geom_smooth(aes(quality-2, 
                    sulphates),
                method = "lm", 
                se = FALSE,size=2)

```

##### Citric Acid v Quality
The correlation between citric acid and quality was 0.23.  The median value goes
from 0.09 at quality of 4 to 0.42 for a quality value of 8.  The Lower quality 
values have large citric acid outliers while the higher quality wines tend to 
have more similar median and mean values; The box plot makes this much clearer.  
The overall trend is higher citric acid values at higher quality wines.  

```{r message=FALSE, warning=FALSE,echo=FALSE}

with(wines, by(citric.acid, factor(quality), summary))

ggplot(aes(factor(quality), 
            citric.acid), 
        data = wines) +
  geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)+
  geom_smooth(aes(quality-2, 
                    citric.acid),
                method = "lm", 
                se = FALSE,size=2)


# calculate correlation coefficient
cor.test(wines$volatile.acidity, wines$citric.acid) 

```
There is a strong correlation between volatile acidity and citric acid.  Maybe a
more complex relationship between acids in wine? 

#### 7 Multivarate Plots

We can see more clearly the correlations between quality, alcohol and 
volatile.acidity more clearly on one plot.
Higher quality wines have higher alcohol and lower volatile.acidity.

```{r corrplots,message=FALSE, warning=FALSE,echo=FALSE}
ggplot(wines,aes(x = rating,y = alcohol,color = volatile.acidity))+
  geom_point(position='jitter',alpha = 0.5,size = 2)+ scale_color_gradient(low = 'red',high='blue')
```

Lets now add sulphates and replot.  Higher quality wines have higher alcohol 
(x-axis), lower volatile acidity (y-axis) and more sulphates (hue)

```{r corrplots2, message=FALSE, warning=FALSE,echo=FALSE}
ggplot(wines, aes(x=alcohol,y=volatile.acidity)) +
      geom_point(aes(color =log10(sulphates)),alpha=1/2,size=2)+
      scale_colour_gradient2(low='red', high='blue')+
      facet_grid(rating ~ .)

```

#### 8 Building a linear model
Question: Can we use our variables to predict quality?
Using all the variables, linear regression can be attempted to predict the 
quality of wine.  From the summary data, the coefficient of determination 
indicates the fitted line will miss many points.  This is verified from the 
residual data plot vs the fitted line.        

```{r message=FALSE, warning=FALSE,echo=FALSE}
wines_model<-lm(quality ~   fixed.acidity+
                            volatile.acidity+
                            citric.acid+
                            residual.sugar+
                            chlorides+
                            free.sulfur.dioxide+
                            total.sulfur.dioxide+
                            density+
                            pH+
                            sulphates+
                            alcohol,
                            data = wines)
mtable(wines_model)
summary(wines_model)
plot(wines_model,which = c(1:1))

```

P-values for fixed acidity, citric acid, residual sugar and density are greater 
then 5%, this indicates they may not make a significant contribution to our 
hypothesis with 95% confidence.  This contradicts the correlation results for 
citric acid - maybe there is a more complex relationship between citric acid 
and quality and other acidic variables than at first thought.  The R-squared 
value is 0.36 but the p-value is less than 5% indicating that the variables do 
have impact on the quality rating.  The residual plot also shows higher 
deviations from the fitted curve line at lower and higher quality ratings.

A linear model can therefore be summarized as:

quality = 21.965 + 0.025fixed.acidity - 1.084volatile.acidity - 0.183citric.acid
+ 0.016residual.sugar-1.874chlorides + 0.004free.sulfur.dioxide -
0.003total.sulfur.dioxide - 17.881density- 0.414pH + 0.916sulphates +
0.276alcohol
 
##### Predictions
Using real values of the viable we can test our predictive model:

```{r message=FALSE, warning=FALSE}
#using these variables expect a value around 3
predict.lm(wines_model, data.frame( alcohol=9.0, chlorides=0.074, free.sulfur.dioxide=10.0, pH=3.25, sulphates=0.57, total.sulfur.dioxide=47,volatile.acidity=0.58, citric.acid=0.66,residual.sugar=2.2,density=1,
                                    fixed.acidity=11.6), type="r")

#using these variables expect a value of 5
predict.lm(wines_model, data.frame( alcohol=10.3, chlorides=0.069, free.sulfur.dioxide=9.0, pH=3.3, sulphates=1.2, total.sulfur.dioxide=23,volatile.acidity=0.66, citric.acid=0.22,residual.sugar=2.2,density=0.99,
                                    fixed.acidity=7.7), type="r")


#using these variables expect a value of 7
predict.lm(wines_model, data.frame( alcohol=11.5, chlorides=0.083, free.sulfur.dioxide=21.0, pH=3.7, sulphates=0.71, total.sulfur.dioxide=59,volatile.acidity=0.66, citric.acid=0.39,residual.sugar=3.2,density=0.99,
                                    fixed.acidity=9.8), type="r")


```
The model predictions are skewed towards our middle quality rating (5 and 6).
This agrees with the frequency count for the quality variable - a heavy bias 
towards the middle quality rating values.

#### 9 Final plots

Wine Quality is displayed in histogram form as Figure 1.  Values are heavily 
biased towards the median values.  Few wines are categorized as having a wine
quality of 3 or 8.  In total there were 1599 observations in the data set of
these 1319 were rated at a quality of 5 or 6, this accounts for 82% of the wine
quality ratings.  Only 10 and 18 wines were rated at 3 or 8 respectively, only
1.8% of the total.  The median wine scores have a much lighter blue colour on 
the colour heat map, but low and high quality wine quality scores were a much
darker shade of blue.

```{r plot, results= 'hide',message=FALSE,warning=FALSE,fig.width=8,fig.height=5,echo=FALSE}

ggplot(data=wines, aes(x=factor(quality))) + geom_bar(col='red',aes(fill=..count..))+labs(x="Wine Quality Score", y="Count of Score")
length(wines$quality)
length(wines$quality[wines$quality==5])
length(wines$quality[wines$quality==6])
length(wines$quality[wines$quality==3])
length(wines$quality[wines$quality==8])
```

Figure 2 demonstrates the relationship between alcohol, sulphates, volatile 
acidity and citric acid versus wine rating as boxplots.  Wine rating summarises
the wine quality from 3 to 8 as bad, average and good.  Alcohol percent by
volume with a median around 11.5 shows the highest rating of good.  In this
plot bad and average ratings show a simlar alcohol content.  Though average
rated wines had a significant number of outliers for greater alcohol content,
suggesting the rating of these wines were boosted by having higher alcohol
content. Sulphates show a upwards trend of increasing sulphate conent with wine
rating. As witnessed in the alcohol boxplot, the average rating for sulphates 
shows the highest number of outliers with increasing sulphate content.  
A Citric acid vs wine rating shows a general trend upwards with citric acid
content.  Volatile acidity demonstrates a negative relationship with wine 
rating. 

```{r corrplots5, results = 'hide',message=FALSE, warning=FALSE,fig.width=9, fig.height=6,echo=FALSE}

p1<- ggplot(wines, aes(x=rating, y=alcohol,fill=rating)) +
      geom_boxplot(alpha = 0.7)+theme_linedraw()+
      scale_y_continuous(name = 'alcohol % by volume',breaks = seq(8,16,2),limits = c(8,16))+
      ggtitle("Alcohol vs Wine Rating")

p2<- ggplot(wines, aes(x=rating, y=sulphates,fill=rating)) +
      geom_boxplot(alpha = 0.7)+theme_linedraw()+theme(legend.position="none")+
      scale_y_continuous(name = 'sulphates g/dm^3',breaks = seq(0.25,2,0.5),limits = c(0.3,2))+
      ggtitle("Sulphates vs Wine Rating")

p3<- ggplot(wines, aes(x=rating, y=volatile.acidity,fill=rating)) +
      geom_boxplot(alpha = 0.7)+theme_linedraw()+theme(legend.position="none")+
      scale_y_continuous(name = 'volatile.acidity g/dm^3',breaks = seq(0,2,0.5),limits = c(0,2))+
      ggtitle("Volatile acidity vs Wine Rating")


p4<- ggplot(wines, aes(x=rating, y=citric.acid,fill=rating)) +
      geom_boxplot(alpha = 0.7)+theme_linedraw()+theme(legend.position="none")+
      scale_y_continuous(name = 'citric.acid',breaks = seq(0,1,0.2),limits = c(0,1))+
      ggtitle("Citric Acid vs Wine Rating")
      grid.arrange(p1,p2,p3,p4,ncol=2,top = 'Figure 2 Selected variables vs Quality')

```

Figure 3 shows the relationship between alcohol and volatile acidity with wine 
rating.  The 'good' points tend to be clustered at points in the figure with 
higher alcohol and lower volatile acidity values.  The data set in wine quality
terms is unbalanced with many average wines but few with poor or good ratings.

```{r corrplots8, results = 'hide',message=FALSE, warning=FALSE,fig.width=9, fig.height=6,echo=FALSE}

 ggplot(data = wines,aes(x = volatile.acidity, y = alcohol,colour = rating)) +
  geom_point(size=2,alpha = 0.6) + scale_color_brewer(palette="Set1")+
  geom_smooth(method = "lm", se = FALSE,size=1)+
  ggtitle('Figure 3 Alcohol vs Volatile Acidity by Wine Rating')+
  scale_x_continuous(name = 'volatile acidity g/dm^3')+
  scale_y_continuous(name = 'alcohol % volume')
 
```

### 10 Reflection
This tidy data set contains 1,599 red wines with 11 variables on the chemical 
properties of the wine. At least 3 wine experts rated the quality of each wine,
providing a rating between 0 (very bad) and 10 (very excellent).  My goal was to
determine which chemical properties influence the quality of red wines.  Each 
variable was investigated and then the relationships between these variables 
uncovered by correlation.  A linear model was built using the variables and 
comparison was made between the linear model and actual values.

By correlation the strongest chemical properties affecting quality appeared to 
be alcohol, supplicates, citric acid and volatile acidity.  The linear model 
also picked out alcohol, sulphates and volatile acidity as the most likely 
determining factors of quality.  Though the linear model did not demonstrate a 
strong relationship between quality and citric acid.  The narrowness of the 
quality rating values; relatively few at lower and higher ratings meant the 
linear model was biased towards the average rated wines. 

The analysis was carried out from Portuguese wines, it would be interesting to 
see if similar results were found using wines from another country such as Spain
which has a similar geography.

### References
* https://en.wikipedia.org/wiki/Acids_in_wine
* http://winefolly.com/review/understanding-acidity-in-wine/
* http://t-redactyl.io/blog/2016/04/creating-plots-in-r-using-ggplot2-part-10-boxplots.html
* http://docs.ggplot2.org/0.9.3.1/scale_gradient2.html
* http://waterhouse.ucdavis.edu/whats-in-wine/volatile-acidity
* http://blog.minitab.com/blog/adventures-in-statistics/




